Implementing Capon Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging

### 2018-02-24 HRUIonGPU/conf_article.tex
Adding HRUIonGPU article

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Implementing Capon Adaptive Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging}

@@

\begin{abstract}
\boldmath
Recent results \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012} suggest that the Capon beamformer can provide increased lateral resolution when applied in medical ultrasound imaging. In this paper, the high computational complexity of the Capon beamformer is targeted with the use of Graphics Processing Units (GPUs). We also present in-vivo results with Capon beamforming applied on a full cardiac cycle in addition to simulations. With GPUs we are able to process a $70\degree$ sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.

For a typical cardiac ultrasound image of $80\times400$ pixels ($70\degree$ sector, 15 cm range) acquired using a 2.5 MHz, M=64 element phased array, we obtain 10 fps (subarray length L=M/2, temporal smoothing over 3 samples). If we do a 2-element pre-beamforming to get the channel count down to 32, the frame rate increases to 44 fps. For a 32 element phased array we need less beams to cover the sector ($40\times400$ pixels), hence with the same parameters the frame rate increases to 87 fps. After optimization the bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels. This is an $O(L^3)$ operation, and when L is larger than 16 (M = 64), real time performance is not possible for cardiac imaging with current high-end hardware. The target GPU has been Nvidia’s Quadro 6000, capable of 1 Tflops.

@@

\section{Introduction}
Capon adaptive beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}. For instance, some interesting trade-offs in addition to just improved resolution were introduced, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}. Despite these benefits, widespread use has been limited by the immense computational complexity associated with the method \cite{So2011}.  In this paper we introduce the first implementation of the Capon beamformer exhibiting real-time performance when applied in a typical cardiac ultrasound imaging setting. To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPUs).

Previous work on implementing a parallel Capon beamformer has mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Sinha2002}. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011} have implemented a Capon beamformer on the GPU that produced real apodization weights based on real data. Under this constraint the Capon beamformer is not allowed to micro steer the main lobe, and therefore its full potential is not reached. Micro steering of the main lobe is an important property of the Capon beamformer in order for it to improve edge definitions and contrast throughout an ultrasound image. Another aspect of using complex data is that even though the number of aritmetic operations per sample is increased we can now sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum. All results in this paper are reported with the use of baseband data.

@@

\section{Background}
\subsection{Capon Adaptive Beamforming}
The conventional form of array beamforming is delay-and-sum (DAS). Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ at time $n$ is delayed to broadside with an appropriate delay $\Delta_m[n]$. The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}^{\Delta}[n]. \label{eq:z}
\end{align}
In conventional beamforming, the weighting $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).

As shown in Figure \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a weight vector based on the data at hand.  For Capon adaptive beamforming the weights are found that minimize the beamformer output power while maintaining unit gain at broadside. The result is that interference impinging from other directions than the point of focus is suppressed \cite{Synnevag2007}.
%\textbf{(Vi skal implementer dette. vis bare vektene. Dropp ligning 2, siter, bruk R-hat med en gang. Få med den endelige feilen )}

@@

\begin{figure}
\centerline{\includegraphics[width=3in]{gfx/beamforming_mv.pdf}}
\caption{Adaptive beamforming. Impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$. Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a weighted coherent sum.}
\label{fig:mvbf}
\end{figure}

To calculate the Capon beamformer output we need three equations. First the sample covariance matrix has to be calculated,
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n']\vec{x}_l^{\Delta}[n']^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of samples included in time, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$. Note that the dimension of the covariance matrix is $L \times L$. The covariance matrix is usually loaded with a diagonal factor $\epsilon$ to ensure numerical stability, 
\begin{align}\label{eq:diag}
\mat{\hat{R}} = \mat{\breve{R}} + \epsilon\mat{I}.
\end{align}
A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon &= d \times tr\{\mat{\breve{R}}\}, \\
d &= 1/L
\end{align}
has been applied in much of the resent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Mehdizadeh2012}.

Second, we calculate the Capon weight vector as
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} = \frac{\vec{b}}{\vec{a}^H\vec{b}},
\end{align}
where $\vec{a} = [1_0, 1_1, ..., 1_{L-1}]^T = \vec{1}_L$ when $\vec{x}$ is pre-delayed. 
And last, the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights,
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n]. \label{eq:z_mv}
\end{align}

@@

\subsection{GPU Compute Model}
CUDA by Nvidia has been our choice of GPU developer environment \cite{Nvidia2011}. The CUDA compute model is based on execution of a kernel function across a grid of compute threads, where the kernel function describes the work to be done by each thread. As depicted in Figure \ref{fig:gpulayout}, each position in the grid holds a block of threads, with maximal size of 1024 threads, that supports low-cost synchronization if needed. 
%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
From a CUDA perspective the GPU consists of one or more stream multi processors (SMs), where each SM performs single-instruction-multiple-data (SIMD) operations. A group of blocks is scheduled to each SM until all blocks in the grid have been processed.

An SM has available a limited amount of near-core memory and registers. Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory latency. Global reads and writes on the GPU, in addition to CPU-GPU transfers, have high penalty, and should be minimized as much as possible. All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration. In the next section, we will describe how this mapping can be done for the Capon beamformer.

\begin{figure}
\centerline{\includegraphics[width=3in]{gfx/gpu_layout_vertical_lr.png}}
\caption{Calculation of Capon adaptive weights mapped to GPU architecture. A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups. Note that several samples can be processed per block of threads if $L$ is small. The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
\label{fig:gpulayout}
\end{figure}

\section{Parallel Capon}\label{sec:meth}
If we analyze (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the output $z[n]$. As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be further divided into a small set of parallel processes. That is, for each data vector $\vec{x}^\Delta$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
\text{2) } &\text{Solve the linear system of equations } \vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
%\end{array}
\end{align*}

The computational complexity of these three steps is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively.

\subsection{Calculating Multiple Sample Covariance Matrices}
The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K)$. Note that for a typical array configuration ($M \in [32, 128]$) we do more spatial than temporal smoothing. Hence, $N_K$ is usually small compared to the number of subarrays ($N_L$). If we let $L = M/\alpha$, $L^2N_L$ becomes $(\alpha - 1)L^3 + L^2$, $\alpha \in [1, M]$. A common choice for $\alpha$ is between two and four \cite{Synnevag2007, Mehdizadeh2012, Synnevag2007a, Austeng2008}. In this interval, calculating the sample covariance matrix is cubic with $L$. This is actually the same as for inverting an $L \times L$ matrix. However little attention has been devoted to this step in the literature.

From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the other elements. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ as in \cite{Chen2011}. This will minimize global reads and writes per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length we then can process, $L_{max}=32$, in order to avoid block-to-block synchronization.

The cubic complexity with $L$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlaps across subarrays. Element $(i,j)$ of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  + \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}^{\Delta}[n']x_{j+N_L-1}^{\Delta}[n']^* \nonumber \\
 &- x_{i-1}^{\Delta}[n']x_{j-1}^{\Delta}[n']^H) \label{eq:sliding}
\end{align}
%The same is true in time where
%\begin{align}
%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}^{\Delta}_l[n+K]\vec{x}^{\Delta}_l[n+K]^H \nonumber \\
%&- \vec{x}^{\Delta}[n-K-1]\vec{x}^{\Delta}[n-K-1]^H).
%\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
%Both sliding across subarrays and in time can be realized simultaneously if we hold an $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer. However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only. 
The complexity of calculating $\mat{\hat{R}}$ is now reduced to $O(L^2)$. Important to notice is that sliding will break independence in the sliding dimension. The calculation in (\ref{eq:sliding}) is still independent across time, but not along diagonals in $\mat{\hat{R}}$. We have therefore selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$. And for small subarrays, several covariance matrices are calculated per thread block (Figure \ref{fig:gpulayout}). Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be too high to achieve full occupancy of the GPU.

Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated. However, the way we have decided to organize the GPU compute threads per $\mat{\hat{R}}$, the lower triangle can be calculated with little overhead on the GPU. This also means that we do not rely on an upper triangular solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, even if that could have saved some instructions.

@@

\subsection{Solving Multiple Small Linear Systems}
In the literature, when it comes to solving systems of linear equations on the GPU, focus has mainly been on large systems. GPU libraries therefore often lack a comprehensive collection of batched solvers. The reason for this is two-fold, first the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available. Because of this, it has proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every case. It is therefore not given that solving on the GPU is faster than solving on the CPU in general. However, there are two arguments for doing solving on the GPU. First the CPU, which might already runs a lot of algorithms, is offloaded. Second the transfer of several thousand covariance matrices from CPU-side to GPU-side is spared.

For the results presented in this paper, we have used an unreleased GPU implementation of batched Gauss-Jordan (GJ) elimination (by Nvidia) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$.

\subsection{Compute Beamformer Output}
The formula in (\ref{eq:z_mv}) has been implemented with one thread per element-weight ($L$ threads per weight vector). After the solution vector $\vec{b}$ has been placed in near-core memory, one of the $L$ threads finds the sum of $\vec{b}$. Each of the $L$ threads then reads one element of $\vec{b}$ and scales it with the inverse sum just found to get the weight vector $\vec{w}[n]$.

@@

\begin{figure*}[!t]
\centerline{\subfloat[]{\includegraphics[width=3.49in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
\hfil
\subfloat[]{\includegraphics[width=3.49in]{gfx/benchmark_bar_64_80_400.png}%
\label{fig:benchUS64}}}
\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ sector. Execution time is plotted as a function of subarray length $L \in [1, M/2]$. The figures show execution times for the three steps listed in Section \ref{sec:meth}: Calculation of covariance matrices in light gray, solver in gray, and the final beamformer sum in dark gray. For the calculation of covariance matrices, there is one bar per choice of temporal averaging, $K \in [0, 1, 2]$. (a) A 32 element array. (b) A 64 element array.}
\label{fig:bench}
\end{figure*}

\section{Benchmarks and Videos}
Figure \ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on an Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops single precision, 6 GB global memory, and 144 GB/s memory bandwidth). The problem sizes are calculated based on 32 and 64 element arrays with center frequency 2.5 MHz. By assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz. If we image down to a depth of 15 cm we get approximately 400 samples per range line. To correctly sample a $70\degree$ sector with 64 and 32 element arrays ($\lambda/2$ pitch) we need at least 80 and 40 beams respectively. The presented execution times do not include transfer of data from CPU-side to the GPU. What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.

At (cite) one will find videos of a cardiac loop, 
%(add ref to video)
and a simulated phantom 
%(add ref to videos) 
processed with different types of settings given to the Capon beamformer. Cardiac data was acquired using a 64 element and 3.5 MHz phased array. Simulations were done using a 64 element 2.5 MHz phased array.

\section{Discussion}
From Figure \ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, total execution time is 100 ms (10 fps). Taking into account that we use complex data, but less samples per image, this is similar to the performance reported by Chen et al. \cite{Chen2011}. For the 32 element array in Figure (\ref{fig:benchUS32}), execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$). Doubling this number can be interpreted as pre-beamforming a 64 element array down to 32. The frame rate for a 64 element array is then increased from 10 to 44 fps.

>>>>

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Implementing Capon Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging}

@@

\begin{abstract}
\boldmath
Recent work on Capon beamforming suggest that it can provide increased lateral resolution when applied in a medical ultrasound imaging setting. In this paper, the high computational complexity of the Capon beamformer is targeted with the use of a Graphics Processing Unit (GPU). \textsl{In-vivo} results with Capon beamforming applied on a full cardiac cycle in addition to simulations are presented. With the GPU we are able to process a $70\degree$ sector cardiac image from a 64 element phased array at interactive frame rates using both spatial and temporal smoothing.

For a typical cardiac ultrasound image of $80\times400$ pixels ($70\degree$ sector, 15 cm range) acquired using a 2.5 MHz, M=64 element phased array, we obtain 10 fps (subarray length L=M/2, temporal smoothing over 3 samples). If we perform a 2-element pre-beamforming, the channel count is reduced to 32, and frame rate is increased to 44 fps. For a 32 element phased array we need less beams to cover the sector ($40\times400$ pixels), hence with the same parameters the frame rate increases to 87 fps. The target GPU was the Nvidia’s Quadro 6000, capable of 1 Tflops.

%After optimization the bottleneck of the Capon algorithm is to find the inverse covariance matrix for all pixels. This is an $O(L^3)$ operation, and when L is larger than 16 (M = 64), real time performance is not possible for cardiac imaging with current high-end hardware.

@@

\section{Introduction}
Capon beamforming \cite{Capon1969} applied to medical ultrasound imaging has recently shown promising results in several studies \cite{Synnevag2007, Austeng2008, Vignon2008, Viola, Mehdizadeh2012}. Some interesting trade-offs in addition to just improved resolution has been introduced, like maintaining image resolution in combination with higher framerates, smaller probes, or deeper penetration \cite{Synnevag2009}. Despite these benefits, widespread use has been limited by the immense computational complexity associated with the method \cite{So2011}.  In this paper we introduce an implementation of the Capon beamformer aiming for real-time performance when applied in a typical cardiac ultrasound imaging scenario. To accomplish this task we make use of the massive, and easily accessed, parallel processing power found in modern Graphics Processing Units (GPUs).

Previous work on implementing a parallel Capon beamformer has mainly been focused around radar and passive sonar systems using custom made systolic array processors \cite{McWhirter1989, Sinha2002}. For medical ultrasound, Chen et al. \cite{Chen2011a, Chen2011} implemented a Capon beamformer on the GPU that produced real apodization weights based on real data. Under this constraint the Capon beamformer is not allowed to micro-steer the main lobe, and therefore its full potential is not reached. Micro-steering of the main lobe is an important property of the Capon beamformer in order for it to improve structural details and edge definitions throughout an ultrasound image. Another aspect of using complex data is that even though the number of aritmetic operations per sample is increased we can sample at the system bandwidth, keeping the number of samples for a full cardiac imaging sector at a minimum. All results in this paper are reported with the use of baseband data.

@@

\section{Background}
\subsection{Capon Beamforming}
The conventional form of array beamforming in medical ultrasound imaging is delay-and-sum (DAS). Here, the signal $x_m[n]$ recorded at element number $m \in [0,M-1]$ at time $n$ is delayed to broadside with an appropriate delay $\Delta_m[n]$. The beamformer output, $z[n]$, is then the sum of all $M$ channels weighted with a per-element factor $w_m$,
\begin{align}
z[n] = \sum_{m = 0}^{M-1}w_m^*x_m[n - \Delta_m[n]] = \vec{w}^H\vec{x}^{\Delta}[n]. \label{eq:z}
\end{align}
In conventional beamforming, the weighting $\vec{w}$ is usually real, and is applied to trade resolution for a better signal-to-noise ratio (SNR).

As shown in Fig. \ref{fig:mvbf}, the basic idea of an adaptive beamformer is to form a weight vector based on the data at hand.  In Capon beamforming the weights are found that minimize the beamformer output power while maintaining unit gain at broadside. The result is that interference impinging from other directions than the point of focus is suppressed \cite{Synnevag2007}.
%\textbf{(Vi skal implementer dette. vis bare vektene. Dropp ligning 2, siter, bruk R-hat med en gang. Få med den endelige feilen )}

@@

\begin{figure}
\centerline{\includegraphics[width=3.5in]{gfx/beamforming_mv.pdf}}
\caption{Beamforming with adaptive generation of weights. Impinging signal is demodulated, and aligned in phase by applying steering and focusing delays $\Delta_m[n]$. Adaptive weights, $\vec{w}[n]$, are then calculated based on in-phase data, and finally the output is formed by a weighted coherent sum.}
\label{fig:mvbf}
\end{figure}

The Capon beamformer can be summarized in three main steps. First a sample covariance matrix has to be calculated,
\begin{align}
\mat{\breve{R}}[n] = \frac{1}{N_LN_K}\sum_{n'=n-K}^{n+K} \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n']\vec{x}_l^{\Delta}[n']^H,\label{eq:R}
\end{align}
where  $N_K = 2K + 1$ is the number of samples included in time, $N_L = M-L+1$ is the number of subarrays, and $\vec{x}_l^{\Delta}$ is the $l^\text{th}$ subarray $[x_l^{\Delta}[n], \dotso, x_{l+L-1}^{\Delta}[n]]$. Note that the dimension of the covariance matrix is $L \times L$. The covariance matrix is then loaded with a diagonal factor $\epsilon$ to ensure numerical stability, 
\begin{align}\label{eq:diag}
\mat{\hat{R}}[n] = \mat{\breve{R}}[n] + \epsilon\mat{I}.
\end{align}
A weighting proportional to the output power is often used to reduce the need of parameter adjustments, and the trace of $\mat{\breve{R}}$, 
\begin{align}\label{eq:diag_adapt}
\epsilon[n] &= d \times \frac{tr\{\mat{\breve{R}}[n]\}}{L},
\end{align}
has been applied in much of the recent literature on Capon beamforming for medical ultrasound imaging \cite{Synnevag2007, Mehdizadeh2012}.

Second, we solve the linear system of equations 
\begin{align}\label{eq:b}
\vec{b}[n] = \mat{\hat{R}}[n]^{-1}\vec{a},
\end{align}
from which we derive the Capon weight vector as
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\vec{b}[n]}{\vec{a}^H\vec{b}[n]}.
\end{align}
Note that  $\vec{a} = [1_0, 1_1, ..., 1_{L-1}]^T = \vec{1}_L$ since $\vec{x}^\Delta$ is pre-delayed. 
Finally, the beamformer output is calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights,
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n]. \label{eq:z_mv}
\end{align}

@@

\subsection{GPU Compute Model}
CUDA by Nvidia has been our choice of GPU developing environment \cite{Nvidia2011}. The CUDA compute model is based on execution of a kernel function across a large set of threads, where the kernel function describes the work to be done in each thread. As depicted in Fig. \ref{fig:gpulayout}, threads are organized in a grid of thread-blocks, with maximal size of 1024 threads per block. Threads inside a block supports low-cost synchronization if needed. 
%Each block is further divided into groups of 32 threads, known as a warp. The maximum number of threads inside a block is restricted to 512, hence 16 warps. 
From a CUDA perspective the GPU consists of one or more streaming multi processors (SMs), where each SM performs single-instruction-multiple-data (SIMD) operations. A group of blocks is scheduled to each SM until all blocks have been processed.

An SM has a limited amount of near-core memory and registers. Each thread therefore has to be as light on memory as possible and has to perform enough instructions per used byte in order to hide memory and instruction latency. Global reads and writes on the GPU, in addition to CPU-GPU transfers, have high penalty, and should be minimized as much as possible. All of these constraints combined pose a real challenge as the target problem has to map on to the parallel framework both on a grid, block and memory consumption level in order to benefits from parallel acceleration. In the next section, we will describe how this mapping can be performed for the Capon beamformer.

\begin{figure}
\centerline{\includegraphics[width=3.5in]{gfx/gpu_layout_vertical_lr.png}}
\caption{Calculation of Capon adaptive weights mapped to GPU architecture. A recorded echo across the array for a given range and angle (selected cell) is assigned to a group of $L$ threads that can run independently of all other groups.} %Note that several samples can be processed per block of threads if $L$ is small. The task of these $L$ threads is to calculate the Capon beamformer output in the three depicted steps.}
\label{fig:gpulayout}
\end{figure}

\section{Parallel Capon}\label{sec:meth}
If we analyze (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the output $z[n]$. As shown in Fig. \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image at sample level.  Further, based on memory dependencies and to hold memory consumption per thread at a minimum, a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that, to some extent, can be further divided into a small set of parallel processes. That is, for each data vector $\vec{x}^\Delta$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}.\\
\text{2) } &\text{Solve the linear system of equations presented in (\ref{eq:b})}.\\
\text{3) } &\text{Calculate } \vec{w} \text{ as in (\ref{eq:w}) and } z \text{ as in (\ref{eq:z_mv}).}
%\end{array}
\end{align*}

The computational complexity of these three steps is $O(L^2)$ (with optimization), $O(L^3)$, and $O(L)$, respectively.

\subsection{Calculating Multiple Sample Covariance Matrices}
%The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K)$. 
%Note that for a typical array configuration ($M \in [32, 128]$) we do more spatial than temporal smoothing. Hence, $N_K$ is usually small compared to the number of subarrays ($N_L$). If we let $L = M/\alpha$, $L^2N_L$ becomes $(\alpha - 1)L^3 + L^2$, $\alpha \in [1, M]$. A common choice for $\alpha$ is between two and four \cite{Synnevag2007, Mehdizadeh2012, Synnevag2007a, Austeng2008}. In this interval, calculating the sample covariance matrix is cubic with $L$.

From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the other elements. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$ as in \cite{Chen2011}. This will minimize global reads and writes per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length one then can process, $L_{max}=32$, in order to avoid block-to-block synchronization.

The straightforward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$. %O((\alpha-1)L^3+L^2)$ ($\alpha = M/L$). 
This is the same as for inverting an $L \times L$ matrix. In this work we therefore focus on optimizing this step in order to reduce the total complexity. The cubic complexity with $L$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlaps across subarrays. Element $(i,j)$ of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j \in [1, L-1]$ as
\begin{align}
\mat{\breve{R}}_{i,j}[n] &=  \mat{\breve{R}}_{i-1,j-1}[n]  + \sum_{n'=n-K}^{n+K} (x_{i+N_L-1}^{\Delta}[n']x_{j+N_L-1}^{\Delta}[n']^* \nonumber \\
 &- x_{i-1}^{\Delta}[n']x_{j-1}^{\Delta}[n']^H) \label{eq:sliding}
\end{align}
%The same is true in time where
%\begin{align}
%\mat{\breve{R}}[n] &= \mat{\breve{R}}[n-1] + \sum_{l=0}^{N_L-1} (\vec{x}^{\Delta}_l[n+K]\vec{x}^{\Delta}_l[n+K]^H \nonumber \\
%&- \vec{x}^{\Delta}[n-K-1]\vec{x}^{\Delta}[n-K-1]^H).
%\end{align}
%- \mat{\breve{R}}[n-K-1] + \mat{\breve{R}}[n-K].
%Both sliding across subarrays and in time can be realized simultaneously if we hold an $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from this buffer. However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only. 
The complexity of calculating $\mat{\hat{R}}$ is now reduced to $O(L^2)$. It is important to notice that sliding will break independence along the sliding dimension. The calculation in (\ref{eq:sliding}) is still independent across time, but not along diagonals in $\mat{\hat{R}}$. We have therefore selected to use one thread per diagonal in the upper triangle of $\mat{\hat{R}}$. And for small subarrays, several covariance matrices are calculated per thread block (Fig. \ref{fig:gpulayout}). Maximum supported $L$ is with this approach increased to 512.%, however for small $L$ high $M$ configurations, memory consumption per thread will be too high to achieve full occupancy of the GPU.

Taking symmetry into account, only the upper half of $\mat{\hat{R}}$ needs to be calculated. However, the way we have decided to organize the GPU compute threads per $\mat{\hat{R}}$, the lower triangle can be calculated with little overhead on the GPU. This also means that we do not rely on an upper triangular solver to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$.%, even if that could have saved some instructions.

@@

\subsection{Solving Multiple Small Linear Systems}
In the literature, when it comes to solving systems of linear equations on the GPU, focus has mainly been on large systems. GPU libraries therefore often lack a comprehensive collection of batched solvers. The reason for this is two-fold. First the GPU needs to solve thousands of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available. 
%Because of this, it has proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every case. 
It is therefore not necessarily true that solving on the GPU is faster than solving on the CPU. However, there are two arguments for doing solving on the GPU. First the CPU, which might already runs a lot of algorithms, is offloaded. Second the transfer of several thousand covariance matrices from CPU-side to GPU-side is avoided.

For the results presented in this paper, we have used a GPU implementation of batched Gauss-Jordan (GJ) elimination by Nvidia to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$ ($L_{max} = 76$).

\subsection{Compute Beamformer Output}
The formula in (\ref{eq:z_mv}) has been implemented with one thread per element-weight ($L$ threads per weight vector). After the solution vector $\vec{b}[n]$ has been placed in near-core memory, one of the $L$ threads finds the sum of $\vec{b}[n]$. Each of the $L$ threads then reads one element of $\vec{b}[n]$ and scales it with the inverse sum just found to get the weight vector $\vec{w}[n]$.

@@

\begin{figure*}[!t]
\centerline{\subfloat[]{\includegraphics[width=3.7in]{gfx/benchmark_bar_32_40_400.png} \label{fig:benchUS32}}
\hfil
\subfloat[]{\includegraphics[width=3.7in]{gfx/benchmark_bar_64_80_400.png}%
\label{fig:benchUS64}}}
\caption{Benchmark of GPU Capon beamforming for a cardiac image covering a $70\degree$ sector. Execution time is plotted as a function of subarray length $L \in [1, M/2]$. The figures show execution times for the three steps listed in Section \ref{sec:meth}. (a) A 32 element array. (b) A 64 element array.}
\label{fig:bench}
\end{figure*}

\section{Benchmarks and Videos}
Fig. \ref{fig:bench} shows the results from benchmarking the implementation described in Section \ref{sec:meth} on an Nvidia Quadro 6000 graphics card (14 SMs, 1 TFlops single precision, 6 GB global memory, and 144 GB/s memory bandwidth). The problem sizes are calculated based on 32 and 64 element arrays with center frequency 2.5 MHz. By assuming a system bandwidth of 80\% we need a complex sampling frequency of 2 MHz. If we image down to a depth of 15 cm we get approximately 400 samples per range line. To correctly sample a $70\degree$ sector with 64 and 32 element arrays ($\lambda/2$ pitch), we need at least 80 and 40 beams, respectively. The presented execution times do not include transfer of data from CPU-side to the GPU. What is included are all calculations, memory transfers from global to near-core GPU memory, and near core memory access.

There are two videos attached to this paper. One cardiac loop, 
%(add ref to video)
and one simulation, both 
%(add ref to videos) 
processed with the following Capon beamformer settings: $L = 32$, $K = 2$, $d=0.01$. These settings have previously been shown to provide high resolution while maintaining DAS-like speckle \cite{Synnevag2007a} in ultrasound images. Cardiac data was acquired using a 64 element and 3.5 MHz phased array. Simulations were done using a 64 element 2.5 MHz phased array. For the simulation, 16x lateral over-sampling has been used, in order to avoid under sampling of the two bright point scatterers. The cardiac recording only has $30\%$ lateral over sampling.

\section{Discussion}
From Fig. \ref{fig:benchUS64} we see that for $L = M/2 = 32$ and $K=1$, total execution time is 100 ms (10 fps). Taking into account that we use complex data, but less samples per image, this is similar to the performance reported by Chen et al. \cite{Chen2011}. For the 32 element array in Fig. (\ref{fig:benchUS32}), execution time is reduced to 11.5 ms (87 fps) using the same settings ($L=M/2, K = 1$). Doubling this number can be interpreted as pre-beamforming a 64 element array down to 32. The frame rate for a 64 element array is then increased from 10 to 44 fps. Note that lateral oversampling will increase the total execution time with a factor equal the amount of oversampling.
